---
title: "SR-Aging_brain-plots"
author: "Cooper Joseph Sharp"
date: "2025-12-19"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Purpose and note on circularity

These plots are **descriptive** summaries of values extracted from clusters identified in whole-brain analyses.
Because the cluster definition and the extraction come from the same dataset/effect, the relationships shown here are **selection-biased (circular)** and should not be interpreted as independent effect size estimates.

## 1. Load libraries

```{r libraries}
packages <- c(
  "dplyr",
  "ggplot2",
  "readr",
  "tibble"
)

# Install missing packages (comment out if you prefer not to install during knitting)
installed <- rownames(installed.packages())
for (pkg in packages) {
  if (!pkg %in% installed) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

## 2. Data import and preparation

```{r import-data}
# Read the combined CSV file
# (If you knit from a different directory, consider using an absolute path or `here::here()`.)
df <- readr::read_csv("extractions-retest-copy.csv", show_col_types = FALSE)

# Quick sanity checks
stopifnot(is.data.frame(df))
```

### 2.1 Helper utilities (weights + plotting)

```{r helpers}
# Inverse-variance weights with a floor to prevent infinite / extreme weights
make_invvar_weights <- function(var_vec, var_floor = 1e-6) {
  1 / pmax(var_vec, var_floor)
}

# Print a compact weight diagnostic (useful when plots "don't look right")
summarize_weights <- function(w) {
  tibble::tibble(
    n = length(w),
    n_nonfinite = sum(!is.finite(w)),
    min = min(w, na.rm = TRUE),
    p01 = as.numeric(stats::quantile(w, 0.01, na.rm = TRUE)),
    p05 = as.numeric(stats::quantile(w, 0.05, na.rm = TRUE)),
    median = stats::median(w, na.rm = TRUE),
    p95 = as.numeric(stats::quantile(w, 0.95, na.rm = TRUE)),
    p99 = as.numeric(stats::quantile(w, 0.99, na.rm = TRUE)),
    max = max(w, na.rm = TRUE)
  )
}

# Partial residual (added-variable style) for age_demeaned from a fitted lm()
partial_residual_age <- function(model, df, x = "age_demeaned") {
  b <- coef(model)[[x]]
  stats::residuals(model) + b * df[[x]]
}

# Raw scatter: unweighted fit
plot_scatter_raw_unweighted <- function(df, outcome_col, outcome_label) {
  ggplot(df, aes(x = age_demeaned, y = .data[[outcome_col]])) +
    geom_point(alpha = 0.55) +
    geom_smooth(method = "lm", se = TRUE) +
    theme_minimal() +
    labs(
      x = "age_demeaned",
      y = outcome_label,
      title = paste0("Raw scatter (unweighted): ", outcome_label),
      caption = "Selection-biased (circular) visualization; do not treat as an independent effect size."
    )
}

# Raw scatter: weighted fit + point size reflects weights
plot_scatter_raw_weighted <- function(df, outcome_col, weight_col, outcome_label) {
  ggplot(df, aes(x = age_demeaned, y = .data[[outcome_col]])) +
    geom_point(aes(size = sqrt(.data[[weight_col]])), alpha = 0.55) +
    geom_smooth(aes(weight = .data[[weight_col]]), method = "lm", se = TRUE) +
    scale_size_continuous(guide = "none") +
    theme_minimal() +
    labs(
      x = "age_demeaned",
      y = outcome_label,
      title = paste0("Raw scatter (weighted): ", outcome_label),
      caption = "Point size reflects inverse-variance weight; smooth uses the same weights. Selection-biased (circular)."
    )
}

# Added-variable (partial residual) scatter: unweighted
plot_added_variable_unweighted <- function(model_unw, df, outcome_label) {
  pr <- partial_residual_age(model_unw, df)

  ggplot(tibble::tibble(age_demeaned = df$age_demeaned, pr = pr),
         aes(x = age_demeaned, y = pr)) +
    geom_point(alpha = 0.55) +
    geom_smooth(method = "lm", se = TRUE) +
    theme_minimal() +
    labs(
      x = "age_demeaned",
      y = paste0("Partial residuals: ", outcome_label),
      title = paste0("Added-variable plot (unweighted): ", outcome_label),
      caption = "Partial residuals are computed from the unweighted model. Selection-biased (circular)."
    )
}

# Added-variable (partial residual) scatter: weighted
plot_added_variable_weighted <- function(model_w, df, weight_col, outcome_label) {
  pr <- partial_residual_age(model_w, df)

  ggplot(tibble::tibble(age_demeaned = df$age_demeaned, pr = pr, w = df[[weight_col]]),
         aes(x = age_demeaned, y = pr)) +
    geom_point(aes(size = sqrt(w)), alpha = 0.55) +
    geom_smooth(aes(weight = w), method = "lm", se = TRUE) +
    scale_size_continuous(guide = "none") +
    theme_minimal() +
    labs(
      x = "age_demeaned",
      y = paste0("Partial residuals: ", outcome_label),
      title = paste0("Added-variable plot (weighted): ", outcome_label),
      caption = "Partial residuals and smooth correspond to the weighted model. Selection-biased (circular)."
    )
}


# Correlation heatmap for a set of variables (coerces factors to numeric codes)
plot_corr_heatmap <- function(df, cols, title = "Correlation heatmap") {
  dat <- df[, cols, drop = FALSE]

  # Coerce non-numeric to numeric (factors -> integer codes; characters -> factor -> codes)
  for (nm in names(dat)) {
    if (!is.numeric(dat[[nm]])) {
      dat[[nm]] <- as.numeric(as.factor(dat[[nm]]))
    }
  }

  cm <- stats::cor(dat, use = "pairwise.complete.obs")
  long <- as.data.frame(as.table(cm))
  names(long) <- c("Var1", "Var2", "r")

  ggplot(long, aes(x = Var1, y = Var2, fill = r)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
    coord_fixed() +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = title, x = NULL, y = NULL, fill = "r")
}

# ---- Sextile (6-bin) summaries + bar plots ----

# Weighted mean / SE for descriptive summaries.
# Uses an effective sample size (n_eff) and an unbiased-ish weighted variance.
wtd_mean_se <- function(y, w) {
  ok <- is.finite(y) & is.finite(w) & (w > 0)
  y <- y[ok]; w <- w[ok]

  if (length(y) == 0) return(list(mean = NA_real_, se = NA_real_, n = 0))

  mu <- sum(w * y) / sum(w)
  sw <- sum(w)
  sw2 <- sum(w^2)
  n_eff <- if (sw2 > 0) (sw^2 / sw2) else length(y)

  # Denominator for an unbiased-ish weighted variance
  den <- sw - (sw2 / sw)
  v <- if (den > 0) sum(w * (y - mu)^2) / den else NA_real_
  se <- if (is.finite(v) && n_eff > 1) sqrt(v / n_eff) else NA_real_

  list(mean = mu, se = se, n = length(y))
}

summarize_sextiles <- function(df, y, series_label, w = NULL, x = "age_demeaned", n_bins = 6) {
  tmp <- tibble::tibble(x = df[[x]], y = y)
  tmp <- tmp %>%
    dplyr::filter(is.finite(x), is.finite(y)) %>%
    dplyr::mutate(bin = dplyr::ntile(x, n_bins))

  if (is.null(w)) {
    tmp %>%
      dplyr::group_by(bin) %>%
      dplyr::summarise(
        n = dplyr::n(),
        age_min = min(x),
        age_max = max(x),
        x_mean = mean(x),
        y_mean = mean(y),
        y_se = stats::sd(y) / sqrt(n),   # SEM
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        series = series_label,
        bin_label = paste0(round(age_min, 2), " to ", round(age_max, 2))
      )
  } else {
    # Align weights to the same finite rows used above
    ok <- is.finite(df[[x]]) & is.finite(y) & is.finite(w) & (w > 0)
    tmp <- tibble::tibble(x = df[[x]][ok], y = y[ok], w = w[ok]) %>%
      dplyr::mutate(bin = dplyr::ntile(x, n_bins))

    tmp %>%
      dplyr::group_by(bin) %>%
      dplyr::summarise(
        n = dplyr::n(),
        age_min = min(x),
        age_max = max(x),
        x_mean = mean(x),
        y_mean = sum(w * y) / sum(w),
        y_se = sqrt(1 / sum(w)),         # SEM for inverse-variance weighted mean
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        series = series_label,
        bin_label = paste0(round(age_min, 2), " to ", round(age_max, 2))
      )
  }
}


plot_sextile_bars <- function(summary_df, outcome_label, title_prefix) {
  # Expect summary_df to contain: bin (1-6), bin_label, y_mean, y_se
  summary_df$bin <- factor(summary_df$bin, levels = 1:6)

  # Use bin_label for x-axis labels (age ranges within each sextile)
  bin_labels <- summary_df %>%
    dplyr::distinct(bin, bin_label) %>%
    dplyr::arrange(bin)

  ggplot(summary_df, aes(x = bin, y = y_mean)) +
    geom_col(width = 0.8) +
    geom_errorbar(aes(ymin = y_mean - y_se, ymax = y_mean + y_se), width = 0.2) +
    scale_x_discrete(labels = bin_labels$bin_label) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      x = "Age sextile (ntile(age_demeaned, 6))",
      y = outcome_label,
      title = paste0(title_prefix, ": ", outcome_label),
      caption = "Bars show mean within sextile; error bars are ±1 SEM. Selection-biased (circular)."
    )
}

```

## 3. DLPFC NPPI (rew > pun), Friend–Stranger

**Outcome:** `mask.dlpfc_type.nppi_cnum.14_cname.rew.pun_F.S`  
**Variance:** `var_mask.dlpfc_type.nppi_cnum.14_cname.rew.pun_F.S`  
**Main predictor:** `age_demeaned`  
**Covariates:** `sex + avg_tsnr_demeaned + avg_fdmean_demeaned + flip_demeaned + study_demeaned`

```{r dlpfc-model}
# ---- DLPFC model (robust to name-repair on import) ----
# We expect readr::read_csv() may "repair" duplicate column names by appending ...1, ...2, etc.
# This helper finds the single best-matching column name in df.

find_one_col <- function(pattern, df_names = names(df)) {
  hits <- grep(pattern, df_names, value = TRUE)
  if (length(hits) == 1) return(hits)
  if (length(hits) == 0) stop(paste0("No columns match pattern: ", pattern))
  stop(paste0("Multiple columns match pattern: ", pattern, "
", paste(hits, collapse = "
")))
}

# Match either exact name OR name with readr's ...<n> suffix.
# Allow either "." or "-" as the separator because some exports use hyphens.
outcome_dlpfc <- find_one_col("^mask[.-]dlpfc_type[.-]nppi_cnum[.-]14_cname[.-]rew[.-]pun_F[.-]S([.]{3}[0-9]+)?$")
var_dlpfc     <- find_one_col("^var_mask[.-]dlpfc_type[.-]nppi_cnum[.-]14_cname[.-]rew[.-]pun_F[.-]S([.]{3}[0-9]+)?$")
w_dlpfc       <- "weights_dlpfc_type.nppi_cnum.14_cname.rew.pun_F.S"

# Confirm required columns exist (now using detected names)
needed <- c(outcome_dlpfc, var_dlpfc,
            "age_demeaned", "sex", "avg_tsnr_demeaned", "avg_fdmean_demeaned",
            "flip_demeaned", "study_demeaned")
missing <- setdiff(needed, names(df))
if (length(missing) > 0) stop(paste("Missing required columns:", paste(missing, collapse = ", ")))

# If sex is coded numerically but should be categorical, convert to factor.
# (Comment out if you are sure you want a linear numeric effect.)
if (!is.factor(df$sex)) {
  u <- sort(unique(df$sex))
  if (length(u) <= 4) df$sex <- factor(df$sex)
}

# Create inverse-variance weights with a small floor (prevents Inf weights if variance is ~0)
df[[w_dlpfc]] <- make_invvar_weights(df[[var_dlpfc]], var_floor = 1e-6)

# Weight diagnostics (helpful if a few observations dominate)
print(summarize_weights(df[[w_dlpfc]]))
# Fit model EXACTLY as specified, but with the detected outcome column name
fml_dlpfc <- as.formula(sprintf(
  "`%s` ~ age_demeaned + sex + avg_tsnr_demeaned + avg_fdmean_demeaned + flip_demeaned + study_demeaned",
  outcome_dlpfc
))

# Weighted and unweighted versions (for side-by-side plots)
model2_w   <- lm(fml_dlpfc, data = df, weights = df[[w_dlpfc]])
model2_unw <- lm(fml_dlpfc, data = df)

summary(model2_w)
summary(model2_unw)
```

### 3.1 Plots for DLPFC

```{r dlpfc-plots}
label_dlpfc <- "DLPFC NPPI (rew > pun), F–S"

# 0) Correlation heatmap among outcome + predictors (+ variance + weight)
plot_corr_heatmap(
  df,
  cols = c(outcome_dlpfc, "age_demeaned", "sex", "avg_tsnr_demeaned", "avg_fdmean_demeaned",
           "flip_demeaned", "study_demeaned", var_dlpfc, w_dlpfc),
  title = paste0("Correlation heatmap: ", label_dlpfc)
)

# 1) Raw scatter: unweighted vs weighted
plot_scatter_raw_unweighted(df, outcome_dlpfc, label_dlpfc)
plot_scatter_raw_weighted(df, outcome_dlpfc, w_dlpfc, label_dlpfc)

# 2) Added-variable (partial residual) plots: unweighted vs weighted
plot_added_variable_unweighted(model2_unw, df, label_dlpfc)
plot_added_variable_weighted(model2_w, df, w_dlpfc, label_dlpfc)

# 3) Bar plots with age binned into sextiles (6 equal bins)

# 3a) Raw outcome bars (unweighted and weighted shown separately)
sum_raw_unw <- summarize_sextiles(df, y = df[[outcome_dlpfc]], series_label = "Unweighted mean")
plot_sextile_bars(sum_raw_unw, outcome_label = label_dlpfc, title_prefix = "Sextile binned bars (raw, unweighted)")

sum_raw_w <- summarize_sextiles(df, y = df[[outcome_dlpfc]], series_label = "Weighted mean", w = df[[w_dlpfc]])
plot_sextile_bars(sum_raw_w, outcome_label = label_dlpfc, title_prefix = "Sextile binned bars (raw, weighted)")

# 3b) Added-variable outcome (partial residuals differ for weighted vs unweighted models)
pr_unw <- partial_residual_age(model2_unw, df)
pr_w   <- partial_residual_age(model2_w, df)

sum_pr_unw <- summarize_sextiles(df, y = pr_unw, series_label = "Unweighted model PR")
plot_sextile_bars(sum_pr_unw, outcome_label = paste0("Partial residuals: ", label_dlpfc),
                  title_prefix = "Sextile binned bars (added-variable, unweighted)")

sum_pr_w <- summarize_sextiles(df, y = pr_w, series_label = "Weighted model PR", w = df[[w_dlpfc]])
plot_sextile_bars(sum_pr_w, outcome_label = paste0("Partial residuals: ", label_dlpfc),
                  title_prefix = "Sextile binned bars (added-variable, weighted)")
```

## 4. MTG ACT (rew), Friend–Stranger

**Outcome:** `mask.mtg_type.act_cnum.17_cname.rew_F.S.txt`  
**Variance:** `var_mask.mtg_type.act_cnum.17_cname.rew_F.S.txt`  
**Main predictor:** `age_demeaned`  
**Covariates:** `sex + avg_tsnr_demeaned + avg_fdmean_demeaned + flip_demeaned + study_demeaned`

```{r mtg-model}
# ---- MTG model (robust to name-repair on import) ----
outcome_mtg <- find_one_col("^mask[.-]mtg_type[.-]act_cnum[.-]17_cname[.-]rew_F[.-]S[.-]txt([.]{3}[0-9]+)?$")
var_mtg     <- find_one_col("^var_mask[.-]mtg_type[.-]act_cnum[.-]17_cname[.-]rew_F[.-]S[.-]txt([.]{3}[0-9]+)?$")
w_mtg       <- "weights_mtg_type.act_cnum.17_cname.rew_F.S.txt"

needed_mtg <- c(outcome_mtg, var_mtg,
                "age_demeaned", "sex", "avg_tsnr_demeaned", "avg_fdmean_demeaned",
                "flip_demeaned", "study_demeaned")
missing_mtg <- setdiff(needed_mtg, names(df))
if (length(missing_mtg) > 0) stop(paste("Missing required columns:", paste(missing_mtg, collapse = ", ")))

df[[w_mtg]] <- make_invvar_weights(df[[var_mtg]], var_floor = 1e-6)
print(summarize_weights(df[[w_mtg]]))
fml_mtg <- as.formula(sprintf(
  "`%s` ~ age_demeaned + sex + avg_tsnr_demeaned + avg_fdmean_demeaned + flip_demeaned + study_demeaned",
  outcome_mtg
))

model3_w   <- lm(fml_mtg, data = df, weights = df[[w_mtg]])
model3_unw <- lm(fml_mtg, data = df)

summary(model3_w)
summary(model3_unw)
```

### 4.1 Plots for MTG

```{r mtg-plots}
label_mtg <- "MTG ACT (rew), F–S"

# 0) Correlation heatmap among outcome + predictors (+ variance + weight)
plot_corr_heatmap(
  df,
  cols = c(outcome_mtg, "age_demeaned", "sex", "avg_tsnr_demeaned", "avg_fdmean_demeaned",
           "flip_demeaned", "study_demeaned", var_mtg, w_mtg),
  title = paste0("Correlation heatmap: ", label_mtg)
)

# 1) Raw scatter: unweighted vs weighted
plot_scatter_raw_unweighted(df, outcome_mtg, label_mtg)
plot_scatter_raw_weighted(df, outcome_mtg, w_mtg, label_mtg)

# 2) Added-variable (partial residual) plots: unweighted vs weighted
plot_added_variable_unweighted(model3_unw, df, label_mtg)
plot_added_variable_weighted(model3_w, df, w_mtg, label_mtg)

# 3) Bar plots with age binned into sextiles (6 equal bins)

# 3a) Raw outcome bars (unweighted and weighted shown separately)
sum_raw_unw <- summarize_sextiles(df, y = df[[outcome_mtg]], series_label = "Unweighted mean")
plot_sextile_bars(sum_raw_unw, outcome_label = label_mtg, title_prefix = "Sextile binned bars (raw, unweighted)")

sum_raw_w <- summarize_sextiles(df, y = df[[outcome_mtg]], series_label = "Weighted mean", w = df[[w_mtg]])
plot_sextile_bars(sum_raw_w, outcome_label = label_mtg, title_prefix = "Sextile binned bars (raw, weighted)")

# 3b) Added-variable outcome (partial residuals differ for weighted vs unweighted models)
pr_unw <- partial_residual_age(model3_unw, df)
pr_w   <- partial_residual_age(model3_w, df)

sum_pr_unw <- summarize_sextiles(df, y = pr_unw, series_label = "Unweighted model PR")
plot_sextile_bars(sum_pr_unw, outcome_label = paste0("Partial residuals: ", label_mtg),
                  title_prefix = "Sextile binned bars (added-variable, unweighted)")

sum_pr_w <- summarize_sextiles(df, y = pr_w, series_label = "Weighted model PR", w = df[[w_mtg]])
plot_sextile_bars(sum_pr_w, outcome_label = paste0("Partial residuals: ", label_mtg),
                  title_prefix = "Sextile binned bars (added-variable, weighted)")
```

## 5. Session info

```{r session-info}
sessionInfo()
```
